### 2018-09-28 paper.tex
Wip

  We present Guild AI, an open source toolkit that facilitates model
  reuse for application development by applying traditional software
  packaging constructs to the domain of machine learning. Package
  management strategies are central to successful software
  ecosystems. Examples include \href{https://www.npmjs.com/}{npm} for
  JavaScript, \href{https://wiki.debian.org/Apt}{APT} for Debian, and
  \href{https://pip.pypa.io}{pip} for Python. Such tools excel at
  creating traditional software packages. However, machine learning
  applications present unique requirements that call for tool
  specialization. Guild AI combines the proven effectiveness of
  package managers with novel features to enable machine learning
  model reuse for application development across a variety of use
  cases.

@@

Code reuse is a central tenet in software development and is supported
by myriad techniques, abstractions, and tools. Effective code reuse
allows developers to leverage prior work in a consistent, reliable
manner that saves time and improves software quality. In this paper we
consider code reuse in terms of \emph{packaged software} in the
context of machine learning.

\subsection{Traditional packaged software}

Packaged software is software that is organized in a way that can be
easily installed and used by \emph{users}. A user is a consumer of
software, rather than a producer or developer. A user may be software
developer, in the case of consuming software libraries or frameworks,
or she may be an end-user who uses software as a program using a
graphical or command line interface. In both cases, packaged software
may be provided in units that the user can install and start using
quickly and reliably.

@@

\begin{itemize}
\item The Python module \verb|tensorflow| should be installed.
\item Additional Python modules required by \verb|tensorflow| should
  also be installed.
\item Installed software should work as expected on the target system.
\item At a future date \verb|tensorflow| may be similarly uninstalled
  along with its no-longer-needed required packages.
\item The act of installing and uninstalling \verb|tensorflow| should
  not break or otherwise destabilize other installed software on the
  target system.
\end{itemize}

Such a package management facilities are essential to code reuse at
the system level. Package managers allow users to consume and manage
complex software systems with simple commands in a way that maintains
the integrity of other installed software. They support rapid
experimentation with new software as users are free to install, use,
and evaluate packages with the confidence that they can safely remove
that software later.

Package managers promote healthy software ecosystems by facilitating
software \emph{publishing}. Software authors use packaging tools to
create the installable software units---\emph{packages}---that are so
easily consumed by users. By publishing packaged software, developers
signify that their software will work as advertised if installed using
the supported tools.

Immense software ecosystems that are enabled by package managers
include Python, JavaScript, Ruby, Go, and R. The GNU/Linux operating
system itself is an ecosystem of package management ecosystems
including packaging schemes from Debian, RedHat, Gentoo, and Arch
distributions. Package managers are established cornerstones of
software reuse across languages and operating systems.

\subsection{Code reuse and machine learning models}

A \emph{machine learning model} is challenging to define as models
have different contexts. In the most general context, models are
mathematical concepts or language that represents a system (ref). In
the context of machine learning, models may be represented by computer
source code, either explicitly as declared structures, or implicitly
by way of executable instructions that perform model related tasks. A
\emph{trained model} is a model that has learned some representation
of a system through the process of computer based optimization.

For the purpose of our discussion, a machine learning model is a
software representation of a specific mathematical concept that can be
trained by a computer. Models may be in various states:
\emph{uninitialized}, \emph{initialized}, \emph{partially trained},
\emph{fully trained}. Regardless of state, the distinguishing feature
of machine learning models for our purpose is that they support
runnable computer operations.

Models---and in particular \emph{trained models}--- are similar to
traditional software in that they can be executed to perform
computations. In fact, the purpose of training a model is to create a
computer program. The difference between a trained model and a human
authored program is one of method: a trained model is learned by an
algorithm while a human authored program is written manually.

While models may be represented in part by traditional installable
software---e.g. a TensorFlow program that imperatively defines a model
graph structure when run---models require more to be construed as
reusable code: \emph{models must be trainable on novel data}.

The remainder of this paper will outline an approach to package that
incorporates \emph{model operations} and \emph{run management} as a
central enabler of code reuse for machine learning models.

@@

A common approach to sharing and reusing machine learning models is a
\emph{model zoo}---an online repository of reusable models that
includes source code, examples, documentation, and in some cases
pretrained models. The \href{Caffe Model Zoo}{Caffe Model Zoo} was one
of the first model zoos and has inspired several others (ref).

Model zoos encourage code reuse by consolidating related models
(e.g. based on underlying library implementation or application
domain) and providing consistent documentation and repeatable methods
for using the models.

Model zoos, however, do not typically provide tools for packaging,
distributing, and using models, beyond those of publishing source code
and documentation (e.g. GitHub, web sites, etc.)

@@

\href{https://www.tensorflow.org/hub/}{TensorFlow Hub modules} and
\href{https://keras.io/applications/}{Keras Applications} are examples
of framework level facilities that provide a higher degree of model
reuse vis-\'a-vis model zoos. Both TensorFlow Hub modules and Keras
application support tools and patterns for packaging, distributing,
and using model source code and associated pretrained models. They
emphasize the use of \emph{transfer learning} where applicable for
training models on novel datasets.

\subsubsection{Software packages and source code repositories}

Models are commonly reused through traditional software distribution
channels: packages and source code repositories. In both cases, model
source code is provided along with varying levels of documentation and
supporting tools. Unlike model zoos and framework packages, there is
little consistency in quality of code reuse available through these
channels. Developers are left to their own accord to find and use
available models.

@@

As with model zoos, users can discover Guild AI models, including
documentation on their use. Guild AI additionally supports
installation and direct use of models through operations.

Guild AI superficially resembles TensorFlow Hub in that users can
discover, install, and use models quickly and easily. Guild AI however
differs considerably in its implementation. Guild AI does not present
a language level interface but instead defines a set of conventions
that are supported through configuration. Guild AI packaging
facilities can be applied to any machine learning model
implementation.

Guild AI also supports the well established practice of package
maintenance, where a third party can create usable packages without
necessarily coordinating with model developers. This is discussed in
detail in Section \ref{sec:package-maintainer}.

@@

A \emph{project} is a directory that contains model related
code. Model developers maintain projects for their models. Model users
maintain projects for the code that works with models, possibly
training them or otherwise incorporating them into an application.

@@

A Guild AI \emph{package} is a Python wheel distribution (ref) that
was generated using the \verb|guild package| command. Guild AI
packages are no different from standard Python wheel distributions and
are generated using \verb|setuptools|. However, they contain Guild AI
specific entry points that effectively publish the models they
contain.

@@

A \emph{model} in Guild AI is a representation of a machine learning
model that is defined in a Guild file. Models define \emph{operations}
(Section \ref{sec:operations}) that perform model related tasks such
as training or evaluating as well as \emph{resources} (Section
\ref{sec:resources}) that operations may require.

Models may be defined according to the underlying architecture or
intended use of the model. Guild AI does not prescribe any formal
meaning to the term \emph{model} or what operations are supported.

@@

Guild AI does not prescribe any particular set of operations that must
be defined for a model. Model developers are free to name operations
as they see fit, though Guild AI does encourage naming conventions.

Below are some common operation names:

\begin{description}
\item[prepare] prepare a dataset for training
\item[train] train a model from scratch
\item[transfer-learn] train a model using transfer learning
\item[finetune] finetune a trained model
\item[evaluate] evaluate a model
\item[classify] use a trained model to classify inputs
\item[compress] compress a trained model (e.g. using quantization,
  etc.)
\item[export-and-freeze] export a frozen inference
  graph\footnote{TensorFlow specific}
\item[tflite] generate a TF Lite file from a frozen
  model\footnotemark[\value{footnote}]
\end{description}

This list illustrates that model operations can be used to perform a
host of tasks beyond training and evaluation. In fact, well designed
models can encompass all of the operations associated with a machine
learning application development pipeline, from data acquisition and
preparation to application deployment and monitoring.

@@

A Guild AI \emph{resource} represents a set of source files that can
be made available as inputs to a model operation. Resources play a key
role in Guild AI's ability to effective use models. Resources can be
used to resolve project files, download files from URLs, unpack
archives, and reference files generated by other operation.

Listing~\ref{lst:resource} illustrates a resource definition for a
model. The resource \verb|idx-data| defines four source files that
will be made available to any operation that requires that resource.

@@

Runs are described in detail in Section \ref{sec:runs}.

@@

A \emph{run directory} is a directory used for a run. During a run
operation, the run directory is the system current working
directory. Operations should write run artifacts within the current
working directory to ensure that they are associated with the run.

By default, Guild AI generates a new directory for each run and
creates links to required resource sources within that directory. Once
the directory is prepared, Guild AI starts the run operating system
process, which then may access sources within the current working
directory.

Users may optionally specify an alternative run directory for a
run. This is commonly used by model developers during the process of
implementing model operation.s

@@

Guild AI \emph{workflows} are patterns of model use that are enabled
by running operations in specific order. Worflows are not currently
formal constructs in Guild AI but are referenced by convention or
documented usage.

@@

Given a Guild file with a package definition, develolpers can use
Guild AI to generate a Python wheel distribution, which in turn can be
used by users to install package models. Developers can also upload
packages to PyPI to make the package available for search and
installation.

@@

Project models are defined in a project Guild file---i.e. a file named
\verb|guild.yml| that resides in the top level of a projet
directory. Users may run operations on project models by running them
from within the project directory or by refering the project directory
in the operation specified for the \verb|run| command.

@@

It is important to note that model and operation help is automatically
generated from model descriptions. This facility lowers the cost of
model documentation for developers and gives users a consistent
interface for discovering model capabilities and intructions for using
them.

@@

As discussed above, published models---deployed from installed
packages or as projects---can be used by running model
\emph{operations}. Operations generate \emph{artifacts} that can be
used in subsquent operations or as final outputs in an application
development workflow.

@@

The first is the use the \verb|--run-dir| option
to the \verb|guild run| command. This option may be used to specify an
alternative directory for the run. This is helpful when the developer
doesn't want to generate new runs for early stage opertion
development.

@@

In many cases, a developer will already have working model code she
wants to incorporate into a formalized model definition. As with the
role of \emph{packger} (Section~\ref{sec:package-maintainer}), a
developer may even be required to work with upstream sources she
cannot easily change. Guild AI provides a number of features that
support the adaptation of existing model code into usable models.

@@

In cases where a developer cannot modify upstream sources or chooses
not to (e.g. to simplify ongoing mainteance) he may choose to patch
model sources for use in a model. Guild AI supports two methods of
patching.

@@

Configuation inheritance can be applied across Guild files---e.g. to
configuration defined in packages or other projects. This is useful
for extending model definitions from other
developers. Listing~\ref{lst:extend-slim} demonstrates how a model can
inherit all of the operations of another model with a single line of
configuration.

@@

The final method of model definition reuse described is that of
\emph{Python main module reuse}. Main modules---i.e. the modules
referenced in an opertion \verb|main| specification---may reside in
any installed Python package. To reuse the main module, a developer
need only include the package reference in the main specification.

Listing~\ref{lst:tflite_convert} illustates the use of the TensorFlow
module \verb|tflite_convert| to generte a TF Lite file for an
operation.

@@

Listing~\ref{lst:tests} contains two tests: one that verifies model
help---ensuring that what the user sees when she runs
\verb|guild help| is what the developer intendes---and another that
exercises model operations and checks results.

>>>>

  We present Guild AI, an open source toolkit that facilitates model
  reuse for application development by applying traditional software
  packaging constructs to the domain of machine learning. Package
  management strategies are central to successful software
  ecosystems. Examples include \href{https://www.npmjs.com/}{npm} for
  JavaScript, \href{https://wiki.debian.org/Apt}{APT} for Debian, and
  \href{https://pip.pypa.io}{pip} for Python. Such tools excel at
  creating traditional software packages. Machine learning
  applications however present unique requirements that call for tool
  specialization. Guild AI combines the proven effectiveness of
  package managers with novel features to enable machine learning
  model reuse for application development across a variety of use
  cases.

@@

Code reuse is a central tenet in software development and is supported
by myriad techniques, abstractions, and tools. Effective code reuse
allows developers to leverage prior work in a consistent, reliable way
that saves time and improves software quality. In this paper we
consider code reuse in terms of \emph{packaged software} in the
context of machine learning.

\subsection{Traditional Packaged Software}

Packaged software is software that is organized in a way that can be
easily installed and used by \emph{users}. A user is a consumer of
software, rather than a producer or developer. A user may be software
developer, in the case of consuming software libraries or frameworks,
or she may be an end-user who uses software as a program with a
graphical or command line interface. In both cases, packaged software
may be provided in units that the user can install and start using
quickly and reliably.

@@

\begin{itemize}
\item The Python module \verb|tensorflow| should be installed.
\item Additional Python modules required by \verb|tensorflow| should
  also be installed.
\item Installed software should work as expected on the target system.
\item At a future date \verb|tensorflow| may be similarly uninstalled
  along with its no-longer-used required packages.
\item The act of installing and uninstalling \verb|tensorflow| should
  not break or otherwise destabilize other installed software on the
  target system.
\end{itemize}

Such package management facilities are essential to code reuse at the
system level. Package managers allow users to maintain complex
software systems with simple commands in a way that preserves their
integrity. They support rapid experimentation with new software as
users are free to install, use, and evaluate packages with the
confidence that they can safely remove that software later.

Package managers promote healthy software ecosystems by facilitating
software \emph{publishing}. Software authors use packaging tools to
create the installable software units---\emph{packages}---that are so
easily consumed by users. By publishing packaged software, developers
signify that their software will work as advertised if installed using
the applicable tools.

Software ecosystems that are enabled by package managers include
Python, JavaScript, Ruby, Go, and R. The GNU/Linux operating system
itself is an ecosystem of package management ecosystems including
packaging schemes from Debian, RedHat, Gentoo, and Arch
distributions. Package managers are established cornerstones of
software reuse across languages and operating systems.

\subsection{Code Reuse and Machine Learning Models}

A \emph{machine learning model} is challenging to define as models
have different contexts. In the most general context, models are
mathematical concepts or language that represents a system (ref). In
the context of machine learning, models may be represented by computer
source code, either explicitly as declared structures, or implicitly
by way of executable instructions that perform model related tasks. A
\emph{trained model} is a model that has learned some representation
of a system through an automated process of optimization.

For the purpose of our discussion, a machine learning model is a
software representation of a specific mathematical concept that can be
trained by a computer. Models may be in various states:
\emph{uninitialized}, \emph{initialized}, \emph{partially trained},
\emph{fully trained}, etc. Regardless of state, the distinguishing
feature of machine learning models for our purpose is that they
support runnable operations.

Models---and in particular \emph{trained models}---are similar to
traditional software in that they can be executed to perform
computations. In fact, the purpose of training a model is to create a
computer program. The difference between a trained model and a human
authored program is one of method: a trained model is learned by an
algorithm while a human authored program is written manually.

While models may be represented in part by traditional installed
software---e.g. an imperative TensorFlow program that defines a model
graph structure when executed---models require more to be construed as
reusable code: \emph{models must support operations on novel data}. If
they cannot be applied to novel data---either for training or for
inference---they cannot be considered reusable software.

The remainder of this paper will outline an approach to package
management that incorporates \emph{model operations} and \emph{run
  management} as an enabler of code reuse for machine learning models.

@@

A common approach to sharing and reusing machine learning models is a
\emph{model zoo}---an online repository of reusable models that
includes source code, examples, documentation, and in some cases
pretrained models. The Caffe Model Zoo (ref) was one of the first
model zoos and has inspired several others (ref).

Model zoos encourage code reuse by consolidating related
models---e.g. based on underlying library implementation or
application domain---and providing consistent documentation and
repeatable methods for model use.

Model zoos, however, do not typically provide tools for packaging,
distributing, and using models beyond those of publishing source code
and documentation (e.g. GitHub, web sites, etc.)

@@

\href{https://www.tensorflow.org/hub/}{TensorFlow Hub modules} and
\href{https://keras.io/applications/}{Keras Applications} are examples
of framework level facilities that provide a higher degree of model
reuse vis-\'a-vis model zoos. Both TensorFlow Hub modules and Keras
application support tools and patterns for packaging, distributing,
and using model source code and associated pretrained models. They
emphasize the use of \emph{transfer learning} for applying pretrained
models to data from different domains.

\subsubsection{Software Packages and Source Code Repositories}

Models are commonly reused through traditional software distribution
channels: \emph{packages} and \emph{source code repositories}. In both
cases, model source code is provided along with varying levels of
documentation and supporting tools. Unlike model zoos and framework
facilities, the level of code reuse available through these channels
is spotty and inconsistent.

@@

As with model zoos, users can discover Guild AI models, including
documentation. Guild AI additionally supports installation and direct
use of models through operations.

Guild AI superficially resembles TensorFlow Hub in that users can
discover, install, and use models quickly and easily. Guild AI however
differs considerably in its implementation. Guild AI does not use a
programming interface but instead supports its features by way of
declared configuration and system level tools. Guild AI packaging
facilities can be applied to any machine learning model
implementation.

Guild AI also supports the process of \emph{package maintenance},
where a third party---a package maintainer---creates usable packages
without necessarily coordinating with upstream model developers. This
is discussed in Section \ref{sec:package-maintainer}.

@@

A \emph{project} is a directory that contains model related
code. Model \emph{developers} maintain projects for their
models. Model \emph{users} maintain projects for their applications,
which may use models for training, inference, etc.

@@

A Guild AI \emph{package} is a Python wheel distribution (ref) that
was generated using the \verb|guild package| command. Guild AI
packages are generated using \verb|setuptools|. They contain Guild AI
specific metadata that advertises models they contain.

@@

A Guild AI \emph{model} is a representation of a machine learning
model that is defined in a Guild file. Models define \emph{operations}
(Section \ref{sec:operations}) that perform model related tasks such
as training or evaluating as well as \emph{resources} that operations
may require (Section \ref{sec:resources}).

Models may be defined according to the underlying architecture or
intended use of the model.

@@

Guild AI does not prescribe any particular set of operations that must
be defined for a model. Model developers are free to name operations
as they see fit, though Guild AI does encourage the use of consistent
naming conventions.

Below are some common operation names.

\begin{description}
\item[prepare] prepare a dataset for training
\item[train] train a model from scratch
\item[transfer-learn] train a model using transfer learning
\item[finetune] finetune a trained model
\item[evaluate] evaluate/test a model
\item[classify] use a trained model for classification
\item[compress] compress a trained model (e.g. using quantization,
  etc.)
\item[export-and-freeze] export a frozen inference
  graph\footnote{TensorFlow specific}
\item[tflite] generate a TF Lite file from a frozen
  model\footnotemark[\value{footnote}]
\end{description}

This list illustrates that model operations can be used to perform
tasks beyond just training and evaluation. Feature rich models can
encompass all of the operations associated with a machine learning
development pipeline, from data acquisition and preparation through
training and optimization to application deployment and monitoring.

@@

A Guild AI \emph{resource} represents a set of source files that can
be made available as inputs to a model operation. Resources play an
important role in Guild AI's ability to effective use
models. Resources can be used to resolve project files, download files
from URLs, unpack archives, and reference files generated by other
operation.

Listing~\ref{lst:resource} contains a resource definition for a
model. The resource \verb|idx-data| defines four source files that
will be made available to any operation that requires that resource.

@@

Runs are further described in Section \ref{sec:runs}.

@@

A \emph{run directory} is a directory used for a run. When an
operation is run, the run directory is used as the operating system
process working directory. Operations should write run artifacts
within the working directory to ensure they are associated with the
run.

By default, Guild AI generates a new directory for each run and
creates links to required sources within that directory. Once the
directory is prepared, Guild AI starts the run operating system
process, which may read sources within the run directory.

@@

Guild AI \emph{workflows} are patterns of model use that are enabled
by running operations in specific order. Worflows are not currently
formal constructs in Guild AI but are used by convention and may be
implemented by other programs.

@@

Given a Guild file with a package definition, developers can use
Guild AI to generate a Python wheel distribution, which in turn can be
used by users to install package models. Developers can also upload
packages to PyPI to make the package available for search and
installation.

@@

Project models are defined in a project Guild file---i.e. a file named
\verb|guild.yml| that resides in the top level of a project
directory. Users may run operations on project models by running them
from within the project directory or by referring the project directory
in the operation specified for the \verb|run| command.

@@

It is important to note that model and operation help is automatically
generated from model descriptions. This facility lowers the cost of
model documentation for developers and gives users a consistent
interface for discovering model capabilities and instructions for using
them.

@@

As discussed above, published models---deployed from installed
packages or as projects---can be used by running model
\emph{operations}. Operations generate \emph{artifacts} that can be
used in subsequent operations or as final outputs in an application
development workflow.

@@

The first is the use the \verb|--run-dir| option
to the \verb|guild run| command. This option may be used to specify an
alternative directory for the run. This is helpful when the developer
doesn't want to generate new runs for early stage operation
development.

@@

In many cases, a developer will already have working model code she
wants to incorporate into a formalized model definition. As with the
role of \emph{packager} (Section~\ref{sec:package-maintainer}), a
developer may even be required to work with upstream sources she
cannot easily change. Guild AI provides a number of features that
support the adaptation of existing model code into usable models.

@@

In cases where a developer cannot modify upstream sources or chooses
not to (e.g. to simplify ongoing maintenance) he may choose to patch
model sources for use in a model. Guild AI supports two methods of
patching.

@@

Configuration inheritance can be applied across Guild files---e.g. to
configuration defined in packages or other projects. This is useful
for extending model definitions from other
developers. Listing~\ref{lst:extend-slim} demonstrates how a model can
inherit all of the operations of another model with a single line of
configuration.

@@

The final method of model definition reuse described is that of
\emph{Python main module reuse}. Main modules---i.e. the modules
referenced in an operation \verb|main| specification---may reside in
any installed Python package. To reuse the main module, a developer
need only include the package reference in the main specification.

Listing~\ref{lst:tflite_convert} illustrates the use of the TensorFlow
module \verb|tflite_convert| to generate a TF Lite file for an
operation.

@@

Listing~\ref{lst:tests} contains two tests: one that verifies model
help---ensuring that what the user sees when she runs
\verb|guild help| is what the developer intends---and another that
exercises model operations and checks results.
